The Calculus of Linear Constructions

### 2021-10-16 writing/lics22.tex
lics22

  \section{Introduction}
  The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand, and Huet in their landmark work \cite{cc}. In CC types can depend on terms, allowing one to write precise specifications as types. Today, CC and its variations CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean}, and others. These theorem provers have found great success in the fields of software verification, and constructive mathematics.

@@

  Linear Logic is a substructural logic introduced by Girard in his seminal work \cite{girard}. Girard notice that the weakening and contraction rules of Classical Logic when restricted carefully, gives rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} first notice that an analogous restriction to variable usage in simple type theory leads to a linear type theory, where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications. The success of integrating Linear Logic with simple type theory exposes a tantalizing new frontier of integrating linearity with richer type theories.

  Work has been done to extend dependent type theories with linear types. Cervesato and Pfenning extend the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} presents a linear dependent type theory, with syntax and semantics drawing inspiration from DILL \cite{dill}.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's earlier work on mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming in the style of Hoare Type Theory \cite{htt}. Luo et al. introduce the property of essential linearity, and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear, and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

  We present a new linear dependent type system - The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (x \ltype Money\ 5) \rightarrow (Camels + Marlboro)
  \end{equation*}
  This new specification for transaction states that it requires a payment of 5 units of money, the customer is relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

@@

  \noindent \textbf{\textit{Contributions}}: 
  Our contributions can be summarized as follows.
  \begin{itemize}
    \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
    \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation, and subject reduction.
    \item Next, we prove the linearity theorem, showing that the tangible impact linear types have how terms can be structured. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
    \item Additionally, we observe that CLC is highly backwards compatible with CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$. This shows that CLC is consistent.
    \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
    \item Furthermore, we extend CLC with linear inductive data, demonstrating CLC's capabilities for correct by construction imperative programming.
    \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
  \end{itemize}

  \section{The Language of CLC}
  \subsection{Syntax}
  \begin{figure}[h]
    \vspace{-1em}
    \caption{Syntax}
    \vspace{1.5em}
    \begin{grammar}
      <$k$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} predicative levels

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \label{syntax}
  \end{figure}
  The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear, and linear types respectively.
  \begin{figure}[H]
    \vspace{-1em}
    \caption{Correspondence of CLC types and MELL implications}
    \Description{}
    \vspace{-1em}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \vspace{-1em}
    \label{correspondence}
  \end{figure}

  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory are the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic directly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we implicitly limit ! to only be canonically usable. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.

@@

  \subsection{Universes and Cumulativity}
  CLC features two sorts of universes $U$, and $L$ with level indices $*, 0, 1, 2, \cdots$. $U_*$ is the impredicative universe of propositions. $U_k$, and $L_k$ are the predicative universe of non-linear types, and linear types respectively. The main mechanism that CLC uses to distinguish between linear and non-linear types is by checking the universe to which they belong. Basically, terms with types that occur within $U_i$ are unrestricted in their usage. Terms with types that occur within $L_k$ are restricted to being used exactly once.

@@

  \begin{figure}[H]
    \vspace{-1.6em}
    \caption{The Universe Hierarchy}
    \vspace{0.8em}
    \centering
    \begin{tikzcd}
      U_* \arrow[r, ":"', "\preceq", dash] 
      & U_0 \arrow[r, ":"', "\preceq", dash] 
      & U_1 \arrow[r, ":"', "\preceq", dash] 
      & U_2 \arrow[r, ":"', "\preceq", dash] 
      & \cdots \\
      & L_0 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_1 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_2 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & \cdots
    \end{tikzcd}
    \vspace{-1em}
    \label{universe}
  \end{figure}

@@

  Next, we define a $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ relation that merges two mixed contexts $\Gamma_1$, and $\Gamma_2$ into $\Gamma$, by performing contraction on shared non-linear variables. For linear variables, the $\mrg{\_}{\_}{\_}$ relation is defined if and only if each variable occurs uniquely in one context and not the other. This definition of $\mrg{\_}{\_}{\_}$ is what allows contraction for unrestricted variables whilst forbidding it for restricted.

@@

      \inferrule
      { \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        x \notin \Gamma_1 }
      { \mrg{\Gamma_1}
            {\Gamma_2, x \ltype A}
            {\Gamma, x \ltype A} }
      \rname{Merge-L2} 
    \end{mathpar}
    \vspace{-0.8em}
    \label{structural}
  \end{figure}

@@

  \begin{definition} We formally define the terms \textit{non-linear}, \textit{linear}, \textit{unrestricted}, and \textit{restricted}.
    \begin{enumerate}
      \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A : U_i$.
      \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A : L_k$.
      \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if there exists non-linear type $A$, such that $\Gamma \vdash m : A$. A unrestricted term may be used an arbitrary number of times.
      \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if there exists linear type $A$, such that $\Gamma \vdash m : A$. A restricted term must be used exactly once.
    \end{enumerate}
  \end{definition}

@@

  \subsection{Equality and Reduction} \label{reduction}
  The operational semantics, and definitional $\beta$-equality of CLC are presented in Figure \ref{parallel}, all of which are entirely standard.
  \begin{figure}[h]
    \caption{Equality and Parallel Reduction}
    \begin{mathpar}
      \inferrule
      { m_1 \step^* n \\ m_2 \step^* n }
      { m_1 \equiv m_2 : A }
      \rname{Join}

@@

  All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the first author's Github repository. To the best of our knowledge, this is the first machined checked formalization of a linear dependently type theory. We given a hand written version of the proof in the appendix as well.

@@

  \subsubsection{Substitution} \label{subst}
  Though the substitution lemma is considered a boring and bureaucratic result, it is surprisingly hard to design linear typed languages where the substitution lemma is admissible. Much of the difficulty arise during the substitution of arguments containing ! exponential. Perhaps the most famous work detailing the issues of substitution is due to Wadler \cite{substitute}. He defines additional syntax and semantics for the intricate unboxing of ! terms, solving the lack of substitute in Abramsky's term calculus.

@@

  \begin{theorem} 
    Propagation. For any context $\Gamma$, term $m$, type $A$, and sort $s$, if $\Gamma \vdash m : A$ is a valid judgment, then there exists some sort $s$ and level $i$ such that $\overline{\Gamma} \vdash A : s_i$.
    \label{propagation}
  \end{theorem}

  With weakening, substitution, propagation, and various inversion lemmas proven, subject reduction can now be proved by induction on typing derivation.

@@

  \subsubsection{Linearity}
  At this point, we have proven that CLC is type sound. However, we still need to prove that the removal of weakening, and contraction for restricted variables yield tangible impact on the structure of terms. For this purpose, we define a binding aware recursive function $occurs(x, m)$ that counts the number of times variable $x$ appears within term $m$. The linearity theorem asserts that restricted variables are used exactly once within a term, subsuming safe resource usage.

@@

  For cases with branched syntax in the linearity theorem, such as application $(m\ n)$, the $occurs$ function sums up the occurrences of variable $x$ in branches $m$, and $n$. The narity lemma is used to prove that either $occurs(x, m) = 1 \wedge occurs(x, n) = 0$ or $occurs(x, m) = 0 \wedge occurs(x, n) = 1$ is true.

@@

  \subsubsection{Strong Normalization}
  The strong normalization theorem of CLC is proven by construction of a typing, and reduction preserving erasure function to CC$\omega$. We assume familiarity with CC$\omega$ syntax here, and define the erasure function as follows.

>>>>

  \section{Introduction}
  The Calculus of Constructions (CC) is a dependent type theory introduced by Coquand and Huet in their landmark work \cite{cc}. In CC types can depend on terms, allowing one to write precise specifications as types. Today, CC and its variations CIC \cite{cic} and ECC \cite{ecc} lie at the core of popular proof assistants such as Coq \cite{coq}, Agda \cite{agda}, Lean \cite{lean} and others. These theorem provers have found great success in the fields of software verification, and constructive mathematics.

@@

  Linear Logic is a substructural logic introduced by Girard in his seminal work \cite{girard}. Girard notices that the weakening and contraction rules of Classical Logic when restricted carefully, give rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} first notices that an analogous restriction to variable usage in simple type theory leads to a linear type theory, where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications. The success of integrating Linear Logic with simple type theory exposes a tantalizing new frontier of integrating linearity with richer type theories.

  Work has been done to extend dependent type theories with linear types. Cervesato and Pfenning extend the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} presents a linear dependent type theory, with syntax and semantics drawing inspiration from DILL \cite{dill}.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's earlier work on mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming in the style of Hoare Type Theory \cite{htt}. Luo et al. introduce the property of essential linearity and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

  We present a new linear dependent type system --- The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (x \ltype Money\ 5) \rightarrow (Camels + Marlboro)
  \end{equation*}
  This new specification for transaction demands a payment of 5 units of money, and the customer be relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

@@

  \noindent \textbf{\textit{Contributions}}: 
  Our contributions can be summarized as follows.
  \begin{itemize}
    \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
    \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation and subject reduction.
    \item Next, we prove the linearity theorem, showing that the tangible impact linear types have how terms can be structured. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
    \item Additionally, we observe that CLC is highly backwards compatible with CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$. This shows that CLC is consistent.
    \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
    \item Furthermore, we extend CLC with linear inductive data, demonstrating CLC's capabilities for correct by construction imperative programming.
    \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
  \end{itemize}

  \section{The Language of CLC}
  \subsection{Syntax}
  \begin{figure}[h]
    \vspace{-1.4em}
    \caption{Syntax}
    \vspace{1.4em}
    \begin{grammar}
      <$k$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} predicative levels

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \label{syntax}
  \end{figure}
  The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear and linear types respectively.
  \begin{figure}[H]
    \vspace{-1.4em}
    \caption{Correspondence of CLC types and MELL implications}
    \Description{}
    \vspace{-1em}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \vspace{-1.6em}
    \label{correspondence}
  \end{figure}

  A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic directly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we implicitly limit ! to only be canonically usable. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.

@@

  \subsection{Universes and Cumulativity}
  CLC features two sorts of universes $U$ and $L$ with level indices $*, 0, 1, 2, \cdots$. $U_*$ is the impredicative universe of propositions. $U_k$ and $L_k$ are the predicative universes of non-linear types and linear types respectively. The main mechanism that CLC uses to distinguish between linear and non-linear types is by checking the universe to which they belong. Basically, terms with types that occur within $U_i$ are unrestricted in their usage. Terms with types that occur within $L_k$ are restricted to being used exactly once.

@@

  \begin{figure}[H]
    \vspace{-1.6em}
    \caption{The Universe Hierarchy}
    \vspace{0.8em}
    \centering
    \begin{tikzcd}
      U_* \arrow[r, ":"', "\preceq", dash] 
      & U_0 \arrow[r, ":"', "\preceq", dash] 
      & U_1 \arrow[r, ":"', "\preceq", dash] 
      & U_2 \arrow[r, ":"', "\preceq", dash] 
      & \cdots \\
      & L_0 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_1 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_2 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & \cdots
    \end{tikzcd}
    \vspace{-0.5em}
    \label{universe}
  \end{figure}

@@

  Next, we define a $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ relation that merges two mixed contexts $\Gamma_1$, and $\Gamma_2$ into $\Gamma$, by performing contraction on shared non-linear variables. For linear variables, the $\mrg{\_}{\_}{\_}$ relation is defined if and only if each variable occurs uniquely in one context and not the other. This definition of $\mrg{\_}{\_}{\_}$ is what allows contraction for unrestricted variables whilst forbidding it for restricted ones.

@@

      \inferrule
      { \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        x \notin \Gamma_1 }
      { \mrg{\Gamma_1}
            {\Gamma_2, x \ltype A}
            {\Gamma, x \ltype A} }
      \rname{Merge-L2} 
    \end{mathpar}
    \vspace{-1em}
    \label{structural}
  \end{figure}

@@

  \begin{definition} We formally define the terms \textit{non-linear}, \textit{linear}, \textit{unrestricted} and \textit{restricted}.
    \begin{enumerate}
      \item A \textit{type} $A$ is \textit{non-linear} under context $\Gamma$ if $\Gamma \vdash A : U_i$.
      \item A \textit{type} $A$ is \textit{linear} under context $\Gamma$ if $\Gamma \vdash A : L_k$.
      \item A \textit{term} $m$ is \textit{unrestricted} under context $\Gamma$ if there exists non-linear type $A$, such that $\Gamma \vdash m : A$. A unrestricted term may be used an arbitrary number of times.
      \item A \textit{term} $m$ is \textit{restricted} under context $\Gamma$ if there exists linear type $A$, such that $\Gamma \vdash m : A$. A restricted term must be used exactly once.
    \end{enumerate}
  \end{definition}

@@

  \subsection{Equality and Reduction} \label{reduction}
  The operational semantics and definitional $\beta$-equality of CLC are presented in Figure \ref{parallel}, all of which are entirely standard.
  \begin{figure}[h]
    \caption{Equality and Parallel Reduction}
    \begin{mathpar}
      \inferrule
      { m_1 \step^* n \\ m_2 \step^* n }
      { m_1 \equiv m_2 : A }
      \rname{Join}

@@

  All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the first author's Github repository. To the best of our knowledge, this is the first machined checked formalization of a linear dependently type theory. We give a hand written version of the proof in the appendix as well.

@@

  \subsubsection{Substitution} \label{subst}
  Though the substitution lemma is regarded as a boring and bureaucratic result, it is surprisingly hard to design linear typed languages where the substitution lemma is admissible. Much of the difficulty arises during the substitution of arguments containing ! exponential. Perhaps the most famous work detailing the issues of substitution is due to Wadler \cite{substitute}. He defines additional syntax and semantics for the intricate unboxing of ! terms, solving the lack of substitute in Abramsky's term calculus.

@@

  \begin{theorem} 
    Propagation. For any context $\Gamma$, term $m$, type $A$, and sort $s$, if $\Gamma \vdash m : A$ is a valid judgment, then there exist some sort $s$ and level $i$ such that $\overline{\Gamma} \vdash A : s_i$.
    \label{propagation}
  \end{theorem}

  With weakening, substitution, propagation and various inversion lemmas proven, subject reduction can now be proved by induction on typing derivation.

@@

  \subsubsection{Linearity}
  At this point, we have proven that CLC is type sound. However, we still need to prove that the removal of weakening and contraction for restricted variables yields tangible impact on the structure of terms. For this purpose, we define a binding aware recursive function $occurs(x, m)$ that counts the number of times variable $x$ appears within term $m$. The linearity theorem asserts that restricted variables are used exactly once within a term, subsuming safe resource usage.

@@

  For cases with branched syntax in the linearity theorem, such as application $(m\ n)$, the $occurs$ function sums up the occurrences of variable $x$ in branches $m$ and $n$. The narity lemma is used to prove that either $occurs(x, m) = 1 \wedge occurs(x, n) = 0$ or $occurs(x, m) = 0 \wedge occurs(x, n) = 1$ is true.

@@

  \subsubsection{Strong Normalization}
  The strong normalization theorem of CLC is proven by construction of a typing and reduction preserving erasure function to CC$\omega$. We assume familiarity with CC$\omega$ syntax here, and define the erasure function as follows.
